Parsing Postfix log files

### 2007-11-12 doc/logparser.tex
Corrections to Coverage, Limitations and Review of Other Parsers.

@@

Full coverage of log lines is easy to achieve yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files (at the time
of writing the parser has 139 rules, fully parsing 93 contiguous log files
from Postfix 2.2 and 2.3), and new rules are easy to add.  Maintaining full
coverage is hard because other servers have different restrictions with
custom messages, RBLs change their messages over time, major releases of
Postfix change warning messages (usually adding more information), etc.\ so
over time the log lines drift and change.  Graph~\refwithpage{rule hits}
shows the number of hits for each rule over all 93 log files; it's obvious
that a small number of rules match the vast majority of the lines, and more
than half the rules match fewer than 100 times.

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (many of the
complications discussed in section~\refwithpage{additional complications}
were detected in this way).  The parser issues warnings when it detects any
errors, some of which may alert the user to a problem, e.g.\ when a queueid
is reused before the previous mail is fully dealt with, when a queueid or
\pid{} is not found,\footnote{There will often be warnings about a missing
queueid or \pid{} in the first few hundred or thousand log lines because the
earlier log lines for those connections or mails are in the previous log
file; loading the saved state from the previous log file will solve this
problem.} or when there are problems tracking a child mail (see
section~\refwithpage{tracking re-injected mail}).  There should be few or
no warnings, and the state table should only contain entries for mails
which had yet to be delivered when the log files ended, or started before
the log files began.

@@

Parser coverage is divided into two topics in this section: log lines
covered, and mails covered.  The former is initially more important, as the
parser must successfully parse every line if it is to be complete, but
subsequently the latter takes precedence because reproducing the path a
mail takes through Postfix is the aim of the parser.  Increasing the
percentage of log lines parsed is relatively simple and not intrusive:
adding new rules or modifying existing rules is simplified by the parser
implementation.  Improving the logical coverage is harder, as the actions
taken by Postfix must be reconstructed, and the new sequence of actions
integrated into the existing model without breaking the existing sequences.
Detecting a deficiency in the parsing algorithm is also significantly
harder than detecting unparsed log lines, as the parser will warn about any
unparsed line, whereas the author must spot a flaw in the parser by
understanding the warnings generated.

@@

    \item Each new Postfix release requires new rules to be written to cope
        with the new log lines.  Similarly new RBLs, new policy servers and
        new administrator defined rejection messages require new rules.

@@

    \item The algorithm does not distinguish between mails where one or
        more mails are rejected and a subsequent mail is accepted; it will
        appear in the database as one mail with lots of rejections followed
        by acceptance.  I don't believe it's possible to make this
        distinction given the data Postfix logs, though it might be
        possible to write a policy server to provide additional logging.

@@

    \item Write the policy server referred to in the limitations above.

@@

@@

    \item [Sawmill Universal Log File Analysis and Reporting] is a general
        purpose commercial product which parses 687 log file formats
        (correct as of 2007/04/15) and produces reports.  Its data
        extraction facilities are quite limited, though it does extract
        three different sets of data for Postfix (as of 2007/04/15, one is
        beta), but they do not appear to be interlinked, nor does it save
        sufficient data for the purposes of this project.  Not attempt is
        made to correlate log lines or deal with the difficulties
        documented in sections~\refwithpage{complications}
        and~\refwithpage{additional complications}.\footnote{If any attempt
        is made there is no reference to it in the documentation available
        on the website.} The source code is available in an obfuscated form
        only (presumably for a fee), and the product is quite expensive, as
        it requires a license for each report which is to be generated.
        The web interface allows creation of dynamic reports based on any
        field, but due to the associated cost the author has not
        experimented with it.

@@

        \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
        Fields extracted: from, to, RBL list, client hostname and client
        IP\@.

        \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
        \newline Fields extracted: from, to, client hostname, client IP,
        relay hostname, relay IP, status, response code, RBL list and
        message id.

@@

        Splunk aims to index all an organisation's logs, providing a
        centralised view capable of searching and correlating diverse log
        sources.  The web interface allows complicated searches, providing
        statistics and graphs in real time.  Saved searches can be
        configured to run periodically and the results emailed to a
        recipient or sent to a shell script, which presumably can publish
        the results as required (though possibly without the graphs and
        detailed statistics).  In the specific case of parsing Postfix
        logs, Splunk extracts some standard fields: to and from addresses,
        HELO, date, host the logs were collected from, and protocol.  It
        parses the standard syslog fields at the beginning of the line, and
        extracts any \texttt{key=<value>} pairs; these are the fields
        listed above.  Searches can be made based on the extracted fields,
        and all text in the line is also available for searching.  Many
        types of reports are available, though all are variations of a bar
        or pie chart, with the exception of bubble and heatmap graphs.  It
        is very easy to drill down through the graphs to extract a portion
        of the data (e.g.\ select the hour with the largest number of
        events, then select a particular host, and finally a specific
        address).  It is not possible to search on partial words, though
        the web interface does provide and refine a list of possible search
        terms as you type.  The searches are quite fast, though as the free
        version has a limit on the amount of data indexed per day the
        volume of data to be searched was merely 45mb; the cheapest
        licensed version costs \$6000, and it still limits the volume of
        data indexed per day.

@@

    \item [Isoqlog] \textit{Isoqlog is an MTA log analysis program written
        in C. It designed to scan qmail, postfix, sendmail and exim logfile
        and produce usage statistics in HTML format for viewing through a
        browser. It produces Top domains output according to Sender,
        Receiver, Total mails and bytes; it keeps your main domain mail
        statistics with regard to Days Top Domain, Top Users values for per
        day, per month and years.\/}

@@

        AWStats will produce simple graphs for many different services, but
        that restricts it to supporting the Lowest Common Denominator.  The
        data it will extract from an MTA log file is limited to: \newline
        \hspace*{2em} time2, email, email\_r, host, host\_r, method, url,
        code and bytesd.  \newline  There does not seem to be an
        explanation of those fields in the documentation.  AWStats has no
        special purpose code to deal with the intricacies of Postfix logs,
        in fact it operates by coercing Postfix logs into
        Apache\footnote{The Apache web server is the most popular HTTP
        server in use for the past 10 years; more information is available
        at \url{http://httpd.apache.org/}} format log files, for analysis
        by its HTTP log file parser.  The converting parser is hard coded,
        hard to extend, and only deals with a small portion of the log
        lines generates by Postfix, but it does correlate log lines by
        queueid.

@@

        YAALA uses a plugin based system to analyse log files and produce
        HTML output reports.  The core code is merely 153 lines long, as
        all the parsing and report generation is handled by modules.  YAALA
        supports storing previously gathered data; it dumps its internal
        data structures using Perl's Storable module~\cite{perl-storable},
        so with some knowledge of the module responsible for parsing
        Postfix log files it should be possible to extract data with an
        external program.  Using YAALA as a base would be only slightly
        less work as both input and output modules would need to be
        written; it may even be more work to implement the parser within
        the constraints of YAALA\@.  YAALA bundles a Postfix parser, which
        uses hard-coded parsing to extract the following fields from
        specific log lines:

        Aggregations: count, bytes (the sum of bytes transferred; count is
        not explained).\newline Keyfields [sic]: date, hour, sender,
        recipient, defer\_count, delay, incoming\_host, outgoing\_host.

        YAALA provides an SQL like mini-language used when generating the
        reports; sample reports can be seen at~\cite{yaala-samples}.
        Example query: \newline \hspace*{2em} \texttt{requests BY file
        WHERE host =\~{} Google}. \newline The mini-language is quite
        limited, and cannot be used to extract data, merely to create
        reports.

@@

        Lire is a general purpose log parser supporting many different
        types of log file.  Its Postfix parser extracts the following data
        from Postfix logs: \textit{The email servers' reports will show you
        the number of deliveries and the volume of email delivered by day,
        the domains from which you receive or send the most emails, the
        relays most used, etc.\/}; notably rejections are not mentioned.
        It supports multiple output formats (text, HTML, PDF and Excel 95).
        Logs can be imported and the extracted data saved, and reports
        generated later.  Accessing this data from another application is
        undocumented; given the source code it should be possible, with
        time, to access the data.

        Unlike other parsers reviewed, Lire does attempt to correlate log
        lines by queueid, but not by \pid{}, so e.g.\ the complete list of
        recipients for a mail should be available.

        Lire does attempt to correlate log lines by queueid, not \pid{},
        but its hard coded parsing extracts only part of the available data
        and makes no attempt to deal with the other complications described
        in sections~\refwithpage{complications} and~\refwithpage{additional
        complications}.  It stores data, but not in an easily accessible
        format.  It generates reports in multiple formats, but those
        reports do not appear to be customisable.

@@

        Fields are correlated based on the queueid (referred to as
        sessionname [sic] within Logrep), but not by \pid{}.  The parsing
        is hard coded, hard to extend, and error prone: empty fields are
        saved when the log line doesn't match the \regex{}, though it
        appears that they will not overwrite existing data.  It also counts
        the number of log lines parsed and skipped.  Most notably
        rejections are completely ignored, making it unsuitable for the
        purposes of this project.  Needless to say it doesn't attempt to
        address any of the complications in
        sections~\refwithpage{complications} and~\refwithpage{additional
        complications} except for correlating by queueid.

        Logrep does not come with any documentation, though some scanty
        documentation is available on its website.  It requires a web
        browser to interact with it, ensuring that automated log processing
        will be difficult.  Sadly all attempts to use Logrep failed, as it
        was unable to access the log files selected; this appears to be a
        bug rather than operator error.  As parsing failed it wasn't
        possible to review the reports Logrep can generate (available in
        HTML only), nor to examine the (undocumented) format in which it
        can save extracted data for subsequent reuse.

@@

While there are other programs available which perform basic Postfix log
parsing (some to a greater level of detail than others), few attempt to
correlate log lines queueid (none correlate by \pid{}) to produce an
overall record of the journey each mail traverses through Postfix.  No
other parser collects the breadth of information gathered by \parsername{},
nor makes it as easy to add new parsing rules.  Most other parsers
immediately generate a report and discard the data extracted from the logs;
those which don't discard the data retain it in a format inaccessible to
other tools.  All of the parsers reviewed can produce a report of greater
or lesser detail and complexity, a facility not offered by \parsername{};
reporting is deferred to a subsequent application.

>>>>

\label{connection reuse}

@@

Full coverage of log lines is easy to achieve yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files (at the time
of writing the parser has 139 rules, fully parsing 93 contiguous log files
from Postfix 2.2 and 2.3), and new rules are easy to add.  Maintaining full
coverage is hard because other servers have different restrictions with
custom messages, \RBL{} messages change over time, major releases of
Postfix change warning messages (usually adding more information), etc.,\
so over time the log lines drift and change.  Graph~\refwithpage{rule hits
graph} shows the number of hits for each rule over all 93 log files; it's
obvious that a small number of rules match the vast majority of the lines,
and more than half the rules match fewer than 100 times.

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (many of the
complications discussed in section~\refwithpage{additional complications}
were detected in this way).  The parser issues warnings when it detects any
errors, some of which may alert the user to a problem, e.g.\ when a queueid
is reused before the previous mail is fully dealt with, when a queueid or
\pid{} is not found,\footnote{There will often be warnings about a missing
queueid or \pid{} in the first few hundred or thousand log lines because
the earlier log lines for those connections or mails are in the previous
log file; loading the saved state from the previous log file will solve
this problem.} or when there are problems tracking a child mail (see
section~\refwithpage{tracking re-injected mail}).  There should be few or
no warnings when parsing, and when finished parsing the state table should
only contain entries for mails which had yet to be delivered when the log
files ended, or started before the log files began.

@@

Parser coverage is divided into two topics in this section: log lines
covered, and mails covered.  The former is initially more important, as the
parser must successfully parse every line if it is to be complete, but
subsequently the latter takes precedence because reproducing the path a
mail takes through Postfix is the aim of the parser.  Increasing the
percentage of log lines parsed is relatively simple and not intrusive:
adding new rules or modifying existing rules is simplified by the
separation of rules and algorithm.  Improving the logical coverage is
harder, as the actions taken by Postfix must be reconstructed, and the new
sequence of actions integrated into the existing model without breaking
the existing sequences.  Detecting a deficiency in the parsing algorithm is
also significantly harder than detecting unparsed log lines, as the parser
will warn about any unparsed line, whereas discovering a flaw in the
parser requires understanding of the warnings produced and the mails
remaining in the state table.  Rectifying a flaw in the parser requires an
understanding of both the parser and Postfix's logs, and investigative work
to determine the cause of the deficiency, followed by further examination
of the logs in developing a solution.

@@

    \item Each new Postfix release requires new rules to be written to cope
        with the new log lines.  Similarly using a new \RBL{}, new policy
        server or new administrator defined rejection messages require new
        rules.

@@

    \item The algorithm does not distinguish between mails where one or
        more mails are rejected and a subsequent mail is accepted; it will
        appear in the database as one mail with lots of rejections followed
        by acceptance (this has already been mentioned in
        section~\refwithpage{connection reuse}).  I don't believe it's
        possible to make this distinction given the data Postfix logs,
        though it might be possible to write a policy server to provide
        additional logging.

@@

    \item Write the policy server referred to in limitation 3 above.

@@

XXX COMPARE AND CONTRAST RATHER THAN REVIEW\@.

XXX HIGHLIGHT THE DIFFERENCE BETWEEN ADDING NEW RULES AND NEW PARSER
FUNCTIONALITY\@.

NEED A BETTER TERM THAN HARD-CODED, THOUGH THAT MIGHT ARISE FROM THE
PREVIOUS TODO ITEM\@.

@@

    \item [Sawmill Universal Log File Analysis and Reporting] \textit{
        Sawmill is a \newline Postfix log analyzer (it also support 686
        other log formats).  It can process log files in Postfix format,
        and generate dynamic statistics from them, analyzing and reporting
        events.  Sawmill can parse Postfix logs, import them into a SQL
        database (or its own built-in database), aggregate them, and
        generate dynamically filtered reports, all through a web interface.
        Sawmill can perform Postfix analysis on any platform, including
        Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other UNIX, and
        more.\/}

        Sawmill is a general purpose commercial product which parses 687
        log file formats (correct as of 2007/04/15) and produces reports.
        Its data extraction facilities are quite limited, though it does
        extract three different sets of data for Postfix (one is beta as of
        2007/04/15), but they do not appear to be interlinked, nor does it
        save sufficient data for the purposes of this project.  No attempt
        is made to correlate log lines or deal with the difficulties
        documented in sections~\refwithpage{complications}
        and~\refwithpage{additional complications}.\footnote{If any attempt
        is made there is no reference to it in the documentation available
        on the website.} The source code is available in an obfuscated form
        only (presumably for a fee), and the product is quite expensive, as
        it requires a license per report which is to be generated.  The web
        interface allows creation of dynamic reports based on any field,
        but due to the associated cost the author has not experimented with
        it.

@@

        \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
        Fields extracted: from, to, \RBL{}, client hostname and client
        IP\@.

        \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
        \newline Fields extracted: from, to, client hostname, client IP,
        relay hostname, relay IP, status, response code, \RBL{} and
        message id.

@@

        Splunk aims to index all an organisation's logs, providing a
        centralised view capable of searching and correlating diverse log
        sources.  The web interface allows complicated searches, providing
        statistics and graphs in real time.  Saved searches can be
        configured to run periodically and the results emailed to a
        recipient or sent to a shell script, which presumably can publish
        the results as required (though possibly without the graphs and
        detailed statistics).  In the specific case of parsing Postfix
        logs, Splunk extracts some standard fields: to and from addresses,
        HELO hostname, date, host the logs were collected from, and
        protocol.  It parses the standard syslog fields at the beginning of
        the line, and extracts any \texttt{key=<value>} pairs; these are
        the fields listed above.  Searches can be made based on the
        extracted fields, and all text in the line is also available for
        searching.  Many types of reports are available, though all are
        variations of a bar or pie chart, with the exception of bubble and
        heatmap graphs.  It is very easy to drill down through the graphs
        to extract a portion of the data (e.g.\ select the hour with the
        largest number of events, then select a particular host, and
        finally a specific address).  It is not possible to search on
        partial words, though the web interface does provide and refine a
        list of possible search terms as you type.  The searches are quite
        fast, though as the free version has a limit on the amount of data
        indexed per day the volume of data to be searched was merely 45mb;
        the cheapest licensed version costs \$6000, and it still limits the
        volume of data indexed per day.

@@

    \item [Isoqlog] \textit{Isoqlog is an MTA log analysis program written
        in C. It designed to scan qmail, postfix, sendmail and exim logfile
        and produce usage statistics in \HTML{} format for viewing through
        a browser. It produces Top domains output according to Sender,
        Receiver, Total mails and bytes; it keeps your main domain mail
        statistics with regard to Days Top Domain, Top Users values for per
        day, per month and years.\/}

@@

        AWStats will produce simple graphs for many different services, but
        that restricts it to supporting the Lowest Common Denominator.  The
        data it will extract from an MTA log file is limited to: \newline
        \hspace*{2em} time2, email, email\_r, host, host\_r, method, url,
        code and bytesd.  \newline  There does not seem to be an
        explanation of those fields in the documentation.  AWStats has no
        special purpose code to deal with the intricacies of Postfix logs,
        in fact it operates by coercing Postfix logs into
        Apache\footnote{The Apache web server is the most popular HTTP
        server in use for the past 10 years; more information is available
        at \url{http://httpd.apache.org/}} format log files, for analysis
        by AWStats' HTTP log file parser.  The converting parser is hard
        coded, hard to extend, and only deals with a small portion of the
        log lines generated by Postfix, but it does correlate log lines by
        queueid.

@@

        YAALA uses a plugin based system to analyse log files and produce
        \HTML{} output reports.  The core code is merely 153 lines long, as
        all the parsing and report generation is handled by modules.  YAALA
        supports storing previously gathered data; it dumps its internal
        data structures using Perl's Storable module~\cite{perl-storable},
        so with some knowledge of the data structure created by the module
        responsible for parsing Postfix log files it should be possible to
        extract data with an external program.  Using YAALA as a base would
        be only slightly less work than starting from scratch, as both
        input and output modules would need to be written; it may even be
        more work to implement the parser within the constraints of
        YAALA\@.  YAALA bundles a Postfix parser, which uses hard-coded
        parsing to extract the following fields from specific log lines:

        Aggregations: count(count is not explained), bytes (the sum of
        bytes transferred).\newline Keyfields [sic]: date, hour, sender,
        recipient, defer\_count, delay, incoming\_host, outgoing\_host.

        YAALA provides an SQL like mini-language used when generating the
        reports; sample reports can be seen at~\cite{yaala-samples}.
        Example query: \newline \hspace*{2em} \texttt{requests BY file
        WHERE host =\~{} Google}. \newline The mini-language is quite
        limited, and cannot be used to extract data for external use,
        merely to create reports.

@@

        Lire is a general purpose log parser supporting many different
        types of log file.  Its Postfix parser extracts the following data
        from Postfix logs: \textit{The email servers' reports will show you
        the number of deliveries and the volume of email delivered by day,
        the domains from which you receive or send the most emails, the
        relays most used, etc.\/}; notably rejections are not mentioned.
        It supports multiple output formats (text, \HTML{}, PDF XXX and
        Excel 95).  Logs can be imported and the extracted data saved, and
        reports generated from it later.  Accessing this data from another
        application is undocumented; given the source code it should be
        possible, with time, to access the data.

        Like AWStats and Logrep, Lire attempts to correlate log lines by
        queueid, but not by \pid{}, so the complete list of recipients for
        a mail should be available; however its hard coded parsing extracts
        only part of the available data and makes no attempt to deal with
        the other complications described in
        sections~\refwithpage{complications} and~\refwithpage{additional
        complications}.  It stores data, but not in an easily accessible
        format, and generates reports in multiple formats, but those
        reports do not appear to be customisable.

@@

        Fields are correlated based on the queueid (referred to as
        sessionname [sic] within Logrep), but not by \pid{}.  The parsing
        is hard coded, hard to extend, and error prone: empty fields are
        saved when the log line doesn't match the \regex{}, though it
        appears that they will not overwrite existing data.  It also counts
        the number of log lines parsed and skipped.  Most notably
        rejections are completely ignored, making it unsuitable for the
        purposes of this project.  It doesn't attempt to address any of the
        complications in sections~\refwithpage{complications}
        and~\refwithpage{additional complications} except for correlating
        by queueid.

        Logrep does not come with any documentation, though some scant
        documentation is available on its website.  It requires a web
        browser to interact with it, ensuring that automated log processing
        will be difficult.  Sadly all attempts to use Logrep failed, as it
        was unable to access the log files selected; this appears to be a
        bug rather than operator error.  As parsing failed it wasn't
        possible to review the reports Logrep can generate (available in
        \HTML{} only), nor to examine the (undocumented) format in which it
        can save extracted data for subsequent reuse.

@@

While there are other programs available which perform basic Postfix log
parsing (some to a greater level of detail than others), few attempt to
correlate log lines by queueid (none correlate by \pid{}) to produce an
overall record of the journey each mail traverses through Postfix.  None of
the reviewed parsers collect the breadth of information gathered by
\parsername{}, nor make it as easy to extend the parser.  Most other
parsers immediately generate a report and discard the data extracted from
the logs; those which don't discard the data retain it in a format
inaccessible to other tools.  All of the parsers reviewed can produce a
report of greater or lesser detail and complexity, a facility not offered
by \parsername{}; reporting is deferred to a subsequent application.
